The role of Similarity in native spoken word recognition

### 2024-07-22 manuscript/manuscript.qmd
New version of introduction

---
title: "The role of Similarity in native spoken word recognition"
---

@@

Listening to non-native speech is costlier than listening to native speech, even for highly proficient bilinguals, and especially in acoustically adverse situations [@lecumberri2010non; @takata1990english]. One of the sources of this increased difficulty is the partial mismatch between the phonology of the native and the non-native languages: some acoustic features embedded in the non-native speech signal do not overlap perfectly with any phonemic category in the listener's native language. For example, imagine a Spanish native with no previous familiarity with any other language listening to French for the first time. This listener may encounter the word /pɔʁt/ (*porte*), translation of door in French. The voiced uvular fricative consonant /ʁ/ and the open-mid back rounded vowel /ɔ/ do not exist in Spanish as phonemic categories. These discrepancies can nonetheless be perceived as allophonic variations the native (Spanish) phonemes /ɾ/ and /o/ [@best2001discrimination], leading to the non-native speech signal engaging lexical processing mechanisms [@weber2004lexical]. Even if unfamiliar with the word-form /pɔʁt/, the Spanish monolingual is likely to activate the phonological representation of its Spanish translation /pweɾta/ (*puerta*), ultimately accessing the right semantic representation. Although the phonetic mismatch between the non-native word and its translation can have a noticeable toll on comprehension [@cutler2004patterns], the fact that word recognition can take place at all in such circumstances illustrates that non-native listeners---even low-proficiency ones---are rarely completely naïve to the language they are listening to.

Imagine now that the same Spanish listener hears English for the first time and encounters the word /dɔː/ (*door*). This time, phonological similarity is of little help: /pweɾta/ and /dɔː/ do not share any phonemes. It would come as a surprise if the listener was able to translate the English word successfully. Non-native listeners rely on the phonological similarity between the word-forms presented in the input and their translations. Luckily for them, most languages share such kind of similarity. This is frequently due to their typological closeness and/or socio-historical events involving the speakers of these languages (e.g., migration, social contact). For example, Romance languages such as Spanish and Catalan share many form-similar translation equivalents (also known as *cognates*) [@schepens2012distributions], as in the case of *puerta* and *porta* (door in Spanish and Catalan, respectively). This type of cross-linguistic similarity has been extremely informative for the study of bilingual lexical processing: words that share high form-similarity with their translations are processed faster and more accurately than words sharing lower phonological similarity. This phenomenon has provided strong evidence in favour of the language-non selective hypothesis of bilingual lexical access, which states that bilinguals activate lexical representations in both languages, even during monolingual situations. This facilitation effect has been reported for comprehension [@thierry2007brain; @midgley2011effects; @dijkstra2010cross], production [@costa2000cognate], learning [@de2000hard; @lotto1998effects; @valente2018does; @elias2022crosslanguage], and translation [@christoffels2006memory]. The role that cross-linguistic phonological similarity might play in monolingual lexical access during non-native speech comprehension is unclear. How much phonological similarity is enough for a monolingual to understand a word from a non-native language? How does phonological similarity interact with other determinants of word recognition?

In the present study, we investigate the extent to which monolinguals can exploit phonological similarity when listening to words in an unfamiliar languages, in order to achieve comprehension, while taking into account other lexically-relevant variables interact with the lexical processing mechanisms involved. We used a translation elicitation task in which participants listened to words from an unfamiliar language, and then tried to guess their translation in their native language. We will henceforth refer to the auditory-presented words heard by participants on each trial as *presented words*, and the correct translation for the presented words as *target translations*. We explored listeners' reliance on phonological similarity to succeed in the task by manipulating the amount of phonological similarity between the presented words and their target translations. Since participants were unfamiliar with the presented language, they could only be able to use word-to-word connections to succeed in the translation of words in an unfamiliar language. We therefore predicted that, if the information provided by the phonological form of the unfamiliar presented words is sufficient for translation, participants' performance should increase when the translation pairs are phonologically more similar.

The extent to which participants can rely on phonological similarity is limited by other factors. In particular, the presence of phonological neighbours (word-forms than differ in one phoneme) may interfere with the activation and selection of the appropriate lexical representation in the native language. Words with denser phonological neighbourhoods are recognised more slowly and less accurately than words with sparser phonological neighbourhoods [@luce1998recognizing; @goldinger1989priming; @luce1990similarity; @hamburger1996phonological; @dufour2003lexical]. This is specially true if such neighbours share higher phonological similarity with the presented word, or are lexically more frequent than the target translation [@luce1998recognizing]. The inhibitory effect of phonological neighbourhoods has also been reported across languages. For instance, @weber2004lexical observed that bilinguals displayed lexical competition effects from words in their native language when performing a word-referent matching task in their L2.

In the case of the present task, the potential effect of cross-language phonological neighbourhood density is critical. The presence of phonologically similar non-translation equivalents (e.g., false friends) may act as distractor during lexical access, obstructing the selection of the appropriate target translation in the listener's lexicon. Even if the presented word and its target translation share high phonological similarity, participants may still provide incorrect translations if the presented word also shares high phonological similarity with words on the target language other than its translation, specially if any of them is more lexically frequent than the target translation. We hypothesised that the size of the facilitatory role of phonological similarity would be inversely proportional to the amount of higher-frequency phonological neighbours of the presented word in the target language. For instance, one could expect English participants to incorrectly translate the Spanish word *botón* as bottom instead of as its correct translation *button* (because the lexical frequency of *bottom* is higher than that of *button* and also phonological closer). To test this prediction, we developed a lexical frequency-dependent measure of cross-language phonological neighbourhood density, in which a neighbour is counted only if it is higher frequency and is one phoneme appart from the presented word. If the phonological neighbourhood density of the target translation affects participants' performance negatively, this would suggest that non-native words trigger competition between words in the native language during translation.

In Experiment 1, we collected data from two groups of British English natives living in the United Kingdom. One group was presented with Catalan words, the other with Spanish words. We examined the extent to which participants were able to use the phonological similarity between the presented word (in Catalan or Spanish) and its target translation to provide accurate responses. In Experiment 2, we tested a group of (European) Spanish natives in the same task, who were presented with Catalan words. Catalan and Spanish are both Romance languages whose close typological distance is reflected in the fact that they share many cognates, where English is a Germanic language that shares considerably fewer cognates with Catalan and Spanish. By testing participants translating words from typologically close and distant languages, we expected to widen the range of the phonological similarity scores of the translation pairs involved in the experimental task, therefore allowing us to explore potential cross-language differences in participant's performance. Participants in Experiments 1 and 2 were surprisingly good at translation some words, which had low phonological similarity with their correct translation. In Experiment 3, we collected additional data from a new group of British English natives, who were presented with Catalan and Spanish words. In contrast with the procedure in Experiment 1, participants in Experiment 3 where asked, after providing a response in each trial, to report whether they had previous knowledge of the meaning of the presented word (e.g., some Spanish words are common knowledge, even for monolingual speakers of English). Finally, we present analyses on the joint data sets of Experiments 1 to 3.

@@

We calculated *Similaritys*, our main predictor of interest, by computing the Levenshtein similarity between the X-SAMPA transcriptions of each pair of translations using the `stringdist` R package [@van2014stringdist]. The Levenshtein distance computes the edit distance between two character strings---in this case, phoneme transcriptions---by counting the number of additions, deletions, and substitutions necessary to make both strings identical [@levenshtein1966binary]. We divided this edit distance by the number of characters included in the longest X-SAMPA transcription of the translation pair. This results in a proportion score, in which values closer to zero correspond to lower Levenshtein distances between phonological transcriptions (i.e., higher similarity), and values closer to 1 correspond to higher Levenshtein distances (i.e., lower similarity). This transformation also help account for the fact that the Levenshtein distance tends to increase with the length of the transcriptions. For interpretability, we subtracted this proportion from one, so that values closer to one correspond to higher similarity between phonological transcriptions, and values closer to zero correspond to lower similarity between phonological transcriptions. For example, the *table* (/teɪbəl/)-*mesa* (/mesa/) translation pair had a `r scales::percent(stringsim(enc2utf8("teɪbəl"), enc2utf8("mesa")))` similarity, while the *train* (`trEIn`)-*tren* (t`r knitr::asis_output("\U027E")`En) translation pair had a `r scales::percent(stringsim(enc2utf8("trEIn"), enc2utf8("tɾEn")))` similarity. Figure 1 summarises the lexical frequency, phonological neighbourhood density and phonological overlap of the words included in the Catalan and the Spanish lists.

@@

After data collection, participants answers were manually coded into the following categories: *Correct*, *Typo*, *Wrong*, *False friend*, *Other*. Responses were coded as *Correct* if the provided string of characters was identical to the orthographic form of the correct translation. Trials were coded as *Typo* if the participant provided a string of characters only one edit distance (addition, deletion, or substitution) apart from the orthographic form of the correct translation (e.g., "pengiun" instead of "penguin"), as long as the response did not correspond to a distinct English word. Responses were coded as *False friend* if the participant provided a phonologically similar incorrect translation. Responses were coded as *Other* (see Data analysis section for more details). Both *Correct* and *Typo* responses were considered as correct, while *Wrong* and *False friend* responses were considered as incorrect. *Other* responses were excluded from data analysis.

Trials in which participants took longer than 10 seconds to respond were also excluded. Participants contributed a total of `r nrow(exp_responses)` valid trials (`r nrow(exp_responses[exp_responses$group %in% c("cat-ENG", "cat-SPA"),])` in Catalan, `r nrow(exp_responses[exp_responses$group %in% c("spa-ENG"),])` in Spanish). The task took approximately 15 minutes to be completed.

@@

To test the practical relevance of each predictor we followed [@kruschke2018bayesian]. We first specified a region of practical equivalence (*ROPE*) around zero ([-0.1, +0.1], in the logit scale). This area indicates the values of the regression coefficients that we considered as equivalent to zero. We then computed the 95% posterior credible intervals (CrI) of the regression coefficients of interest. Finally, we calculated the proportion of the 95% CrI that fell inside the ROPE. This proportion indicates the probability that the true value of the coefficient is equivalent to zero.

All analyses were performed in R environment [@rcoreteam2013language]. We used the tidyverse family of R packages [@wickham2019welcome] to process data and to generate figures. We used the `brms` R package [@burkner2017brms] using the `cmdstanr` back-end to the Stan probabilistic language [@carpenter2017stan] to estimate and compare the models (see Appendix 1 for mode details on the models).

@@

We collected data for a total of `r format(nrow(dataset$cat_eng) + nrow(dataset$spa_eng), big.mark = ",")` trials completed by `r n_participants$cat_eng + n_participants$spa_eng` unique participants. Of those trials, `r format(nrow(dataset$cat_eng), big.mark = ",")` were provided by `r n_participants$cat_eng` unique participants who listened to Catalan words, and `r format(nrow(dataset$spa_eng), big.mark = ",")` trials were provided by `r n_participants$spa_eng` unique participants who listened to Spanish words. We excluded trials in which participants did not enter any text (*n* = `r r_validity$cat_eng["blank"]`), in which a response in a language other than English was provided (e.g., `agua`, *n* = `r r_validity$cat_eng["language"]`), in which participants did not provide a whole word (e.g., `f`, *n* = `r r_validity$cat_eng["incomplete"]`), and in which participants added comments to the experimenter (e.g., `unsure`, (*n* = `r r_validity$cat_eng["other"]`). In addition, we excluded data from participants that self-rated their oral and/or written skills in Catalan and Spanish, or any other second language as four or higher in a five-point scale (*n* = `r excluded_lang`), were diagnosed with a language (*n* = `r excluded_impairment`), or did not contribute more than 80% of valid trials (*n* = `r excluded_ntrials`).

@@

```{r fig-epreds-1}
#| label: fig-epreds-1
#| fig-height: 6
#| fig-width: 10
#| fig-cap: "Posterior model-predicted mean accuracy in Experiment 1. Predictions were generated from 4,000 posterior samples, extracted for different values of *CLPN* (0, 2, 4, 8, 12) and *Similarity* (1-100). Predictions are plotted separately for English participants translating Catalan words, and for English participants translating Spanish words. Lines indicate mean predictions, and intervals indicate 95%, 89%, 78%, 67%, and 50% credible intervals (CrI)."
get_epreds <- function(model, data, n = 1000, 
                       lv = seq(0, 1, length.out = 100), 
                       neigh_n_h = c(0, 2, 4, 8, 12),
                       ...) {

@@

On each trial, participants listened to the audio stimulus by clicking on the play button. For comparability to the PsychoPy version, participants were only allowed to play the audio one time. Participants were explicitly told that they would be only allowed to listen once. The play button vanished after one playthrough. Participants then had to answer three questions based on the audio they had heard on that trial. These questions were presented on the same page, directly below the audio player. They were first asked whether or not they knew the presented word (multiple choice---yes/no). Regardless of their answer on the first question, participants were asked what they thought the translation of the word was in English (or their best guess), and instructed to type their answer in the provided text box. Finally, they were asked to rate how confident they were in their answer on a scale of 0 to 7, where 7 was "very confident" and 0 was "not confident". There was no time limit on the response phase. All questions had to be answered to proceed to the next trial.

>>>>

@@

::: {.content-visible when-format="pdf"}

Learning a new language is a non-trivial task, a large part of which is acquiring a new set of word-forms to correspond to existing semantic representations in their lexicon. Imagine that a Spanish listener hears English for the first time and encounters the word /\textipa{"dO:}/ (*door*). Their knowledge of the phonological form in their native language is of little help: /\textipa{"pweR.ta}/ and /\textipa{"dO:}/ do not share any phonemes. It would come as a surprise if the listener was able to translate the English word successfully without having first learnt the word. However, there are some elements in language that may help the learner in acquiring words in a new language. Many languages share some degree of similarity. This may be due to typological closeness and/or socio-historical events involving the speakers of these languages (e.g., migration, social contact). For example, Romance languages such as Spanish and Catalan share many form-similar translation equivalents (also known as cognates) [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively). Given no prior knowledge of either language, a Spanish native speaker is likely to be much more successful at correctly translating Catalan *porta* than English *door*, due to the phonological and orthographical similarity of the former to the equivalent in their native language. Second language learners have been found to learn cognates more easily than non-cognates [@hall2002automatic; @otwinowska2019more]. This type of cross-linguistic similarity has also been extremely informative for the study of bilingual lexical processing: words that share high form-similarity with their translations are processed faster and more accurately than words sharing lower phonological similarity. This phenomenon has provided strong evidence in favour of the language-non selective hypothesis of bilingual lexical access, which states that bilinguals activate lexical representations in both languages, even during monolingual situations. This facilitation effect has been reported for comprehension [@dijkstra2010cross; @midgley2011effects; @thierry2007brain], production [@costa2000cognate], learning [@de2000hard; @elias2022crosslanguage; @lotto1998effects; @valente2018does], and translation [@christoffels2006memory].

Event counting on this facilitation effect of form-similarity, listening to non-native speech is costlier than listening to native speech, even for highly proficient bilinguals, and especially in acoustically adverse situations [@lecumberri2010non; @takata1990english]. One of potential source of this increased difficulty is the partial mismatch between the phonology of the native and the non-native languages: some acoustic features embedded in the non-native speech signal do not overlap perfectly with any phonemic category in the listener’s native language. For example, imagine a Spanish native with no previous familiarity with any other language listening to French for the first time. This listener may encounter the word /\textipa{"pOKt}/ (*porte*), translation of door in French. When comparing the phonemic transcription to the transcription of the Spanish word /\textipa{"pweR.ta}/ (*puerta*), one may be led to infer that these two words do not have much overlap. But one thing to note is that the voiced uvular fricative consonant /\textipa{K}/ and the open-mid back rounded vowel /\textipa{O}/, which do not exist in Spanish as phonemic categories, can be perceived as allophonic variations the native (Spanish) phonemes /\textipa{R}/ and /o/. A perceptual assimilation model has been proposed to account for how speakers may integrate non-native phonemes into their native phonemic categories based on goodness-of-fit [@best2001discrimination; @best1988examination]. Although phonetic mismatch between the non-native word and its translation can have a noticeable toll on comprehension @cutler2004patterns, the Spanish monolingual may still be able to activate the phonological representation of /\textipa{"pweR.ta}/ (*puerta*) from hearing the word-form /\textipa{"pOKt}/ (*porte*), ultimately allowing them to access the right semantic representation.

:::

::: {.content-visible when-format="docx"}

Learning a new language is a non-trivial task, a large part of which is acquiring a new set of word-forms to correspond to existing semantic representations in their lexicon. Imagine that a Spanish listener hears English for the first time and encounters the word /dɔː/ (*door*). Their knowledge of the phonological form in their native language is of little help: /pweɾta/ and /dɔː/ do not share any phonemes. It would come as a surprise if the listener was able to translate the English word successfully without having first learnt the word. However, there are some elements in language that may help the learner in acquiring words in a new language. Many languages share some degree of similarity. This may be due to typological closeness and/or socio-historical events involving the speakers of these languages (e.g., migration, social contact). For example, Romance languages such as Spanish and Catalan share many form-similar translation equivalents (also known as cognates) [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively). Given no prior knowledge of either language, a Spanish native speaker is likely to be much more successful at correctly translating Catalan *porta* than English *door*, due to the phonological and orthographical similarity of the former to the equivalent in their native language. Second language learners have been found to learn cognates more easily than non-cognates [@hall2002automatic; @otwinowska2019more]. This type of cross-linguistic similarity has also been extremely informative for the study of bilingual lexical processing: words that share high form-similarity with their translations are processed faster and more accurately than words sharing lower phonological similarity. This phenomenon has provided strong evidence in favour of the language-non selective hypothesis of bilingual lexical access, which states that bilinguals activate lexical representations in both languages, even during monolingual situations. This facilitation effect has been reported for comprehension [@dijkstra2010cross; @midgley2011effects; @thierry2007brain], production [@costa2000cognate], learning [@de2000hard; @elias2022crosslanguage; @lotto1998effects; @valente2018does], and translation [@christoffels2006memory].

Event counting on this facilitation effect of form-similarity, listening to non-native speech is costlier than listening to native speech, even for highly proficient bilinguals, and especially in acoustically adverse situations [@lecumberri2010non; @takata1990english]. One of potential source of this increased difficulty is the partial mismatch between the phonology of the native and the non-native languages: some acoustic features embedded in the non-native speech signal do not overlap perfectly with any phonemic category in the listener’s native language. For example, imagine a Spanish native with no previous familiarity with any other language listening to French for the first time. This listener may encounter the word /pɔʁt/ (*porte*), translation of door in French. When comparing the phonemic transcription to the transcription of the Spanish word /pweɾta/ (*puerta*), one may be led to infer that these two words do not have much overlap. But one thing to note is that the voiced uvular fricative consonant /ʁ/ and the open-mid back rounded vowel /ɔ/, which do not exist in Spanish as phonemic categories, can be perceived as allophonic variations the native (Spanish) phonemes /ɾ/ and /o/. A perceptual assimilation model has been proposed to account for how speakers may integrate non-native phonemes into their native phonemic categories based on goodness-of-fit [@best2001discrimination; @best1988examination]. Although phonetic mismatch between the non-native word and its translation can have a noticeable toll on comprehension @cutler2004patterns, the Spanish monolingual may still be able to activate the phonological representation of /pweɾta/ (*puerta*) from hearing the word-form /pɔʁt/ (*porte*), ultimately allowing them to access the right semantic representation.

:::

While phonological similarity across translation equivalents can help with speech recognition, phonological similarity between words with different meanings can hinder the recognition process. In word retrieval, the presence of phonological neighbours (word-forms that differ from each other in one phoneme) may interfere with the activation and selection of the appropriate lexical representation in the native language. Words with denser phonological neighbourhoods are recognised more slowly and less accurately than words with sparser phonological neighbourhoods [@dufour2003lexical; @goldinger1989priming; @hamburger1996phonological; @luce1990similarity; @luce1998recognizing]. This is especially true if such neighbours share higher phonological similarity with the presented word, or are lexically more frequent than the target translation [@luce1998recognizing]. The inhibitory effect of phonological neighbourhoods has also been reported across languages. For instance, @weber2004lexical observed that bilinguals displayed lexical competition effects from words in their native language when performing a word-referent matching task in their L2. The assimilation of non-native phonemes into native categories [@best2001discrimination; @best1988examination] was used to explain how lexical competitors in the native language can be activated even when the target words contain non-native phonemes. In the case of non-native speech recognition, the presence of cross-linguistic pairs which are phonologically similar but differ in meaning (e.g., false friends) may act as distractors during lexical access, obstructing the selection of the appropriate target translation in the listener’s lexicon. For instance, @otwinowska2019more reported that false friends were disadvantaged relative to non-cognates by Polish second language learners, in contrast to cognates which were known better. It is therefore important to investigate the joint effect of both cognates and false friends when investigating the effect cross-linguistic phonological similarity has on word recognition in a foreign language. This could shed light on available strategies and challenges associated with the early stages of second language acquisition.

In the present study, we investigate the extent to which monolinguals can exploit phonological similarity between translation equivalents when listening to words in an unfamiliar language, and how it interacts with competition from phonological neighbours in the native language. We used a translation elicitation task in which participants listened to words from an unfamiliar language, and then tried to guess their translation in their native language. We will henceforth refer to the auditory-presented words heard by participants on each trial as presented words, and the correct translation for the presented words as target translations. We explored listeners’ reliance on phonological similarity to succeed in the task by manipulating the amount of phonological similarity between the presented words and their target translations. Since participants were unfamiliar with the presented language, they could only use phonological similarity between the presented language and their native language to successfully translate the words. We therefore predicted that participants’ performance should increase when the translation pairs are phonologically more similar. We also predicted that there is a minimum threshold of phonological similarity to be sufficient for translation.

We further investigated the effect of phonological competitors. Even if the presented word and its target translation share high phonological similarity, participants may still provide incorrect translations if the presented word also shares high phonological similarity with other words in the native language. The effect is predicted to be especially strong if any of the competitors is more lexically frequent than the target translation. We hypothesised that the size of the facilitatory role of phonological similarity would be inversely proportional to the amount of higher-frequency phonological neighbours of the presented word in the target language. For instance, one could expect English participants to incorrectly translate the Spanish word *botón* as *bottom* instead of as its correct translation *button*, due to combined effect of the close phonological similarity between *botón* and *bottom* along with the high lexical frequency of *bottom* relative to *button*. To test this prediction, we developed a lexical frequency-dependent measure of cross-language phonological neighbourhood density, in which a neighbour is counted only if it is higher frequency and is one phoneme apart from the presented word. If the phonological neighbourhood density of the target translation affects participants’ performance negatively, this would suggest that competitors in the native language affect recognition of non-native words during recognition of foreign speech.We conducted a series of three experiments to test these predictions.

In Experiment 1, we collected data from two groups of British English natives living in the United Kingdom. One group was presented with Catalan words, the other with Spanish words. We examined the extent to which participants were able to use the phonological similarity between the presented word (in Catalan or Spanish) and its target translation to provide accurate responses. In Experiment 2, we tested a group of (European) Spanish natives in the same task, who were presented with Catalan words. Catalan and Spanish are both Romance languages whose close typological distance is reflected in the fact that they share many cognates, where English is a Germanic language that shares considerably fewer cognates with Catalan and Spanish. By testing participants translating words from typologically close or distant languages, we expected to widen the range of the phonological similarity scores of the translation pairs involved in the experimental task, therefore allowing us to explore potential cross-language differences in participant’s performance. One unexpected finding was that participants in Experiments 1 and 2 were surprisingly good at translating a subset of words which had low phonological similarity with their correct translation. We were concerned that this may be caused by some prior knowledge of specific words by our participants, as some Spanish words are commonly seen in media or product labels, making them familiar even to monolingual speakers of English. In Experiment 3, we collected additional data from a new group of British English natives. The design was closely modelled after Experiment 1, except that after providing their response in each trial, participants reported whether they had previous knowledge of the presented word. In the final section of this paper, we present analyses on the joint data sets of Experiments 1 to 3.

@@

We calculated *Similarity*, our main predictor of interest, by computing the Levenshtein similarity between the X-SAMPA transcriptions of each pair of translations using the `stringdist` R package [@van2014stringdist]. The Levenshtein distance computes the edit distance between two character strings---in this case, phoneme transcriptions---by counting the number of additions, deletions, and substitutions necessary to make both strings identical [@levenshtein1966binary]. We divided this edit distance by the number of characters included in the longest X-SAMPA transcription of the translation pair. This results in a proportion score, in which values closer to zero correspond to lower Levenshtein distances between phonological transcriptions (i.e., higher similarity), and values closer to 1 correspond to higher Levenshtein distances (i.e., lower similarity). This transformation also help account for the fact that the Levenshtein distance tends to increase with the length of the transcriptions. For interpretability, we subtracted this proportion from one, so that values closer to one correspond to higher similarity between phonological transcriptions, and values closer to zero correspond to lower similarity between phonological transcriptions. For example, the *table* (`teIb@l`)-*mesa* (`mesa`) translation pair had a `r scales::percent(stringsim(enc2utf8("teɪbəl"), enc2utf8("mesa")))` similarity, while the *train* (`trEIn`)-*tren* (`tREn`) translation pair had a `r scales::percent(stringsim(enc2utf8("trEIn"), enc2utf8("tɾEn")))` similarity. @tbl-stimuli summarises the lexical frequency, phonological neighbourhood density and phonological overlap of the words included in the Catalan and the Spanish lists.

@@

After data collection, participants answers were manually coded into the following categories: *Correct*, *Typo*, *Wrong*, *False friend*, *Other*. Responses were coded as *Correct* if the provided string of characters was identical to the orthographic form of the correct translation. Trials were coded as *Typo* if the participant provided a string of characters only one edit distance (addition, deletion, or substitution) apart from the orthographic form of the correct translation (e.g., "pengiun" instead of "penguin"), as long as the response did not correspond to a distinct English word. Responses were coded as *False friend* if the participant provided a phonologically similar incorrect translation. Responses were coded as *Other* (see Data analysis section for more details). Both *Correct* and *Typo* responses were considered as correct, while *Wrong* and *False friend* responses were considered as incorrect. *Other* responses were excluded from data analysis. Trials in which participants took longer than 10 seconds to respond were also excluded. Participants contributed a total of `r nrow(exp_responses)` valid trials (`r format(nrow(exp_responses[exp_responses$group %in% c("cat-ENG", "cat-SPA"),]), big.mark = ",")` in Catalan, `r format(nrow(exp_responses[exp_responses$group %in% c("spa-ENG"),]), big.mark = ",")` in Spanish). The task took approximately 15 minutes to be completed.

@@

To test the practical relevance of each predictor we followed [@kruschke2018bayesian]. We first specified a region of practical equivalence (*ROPE*) around zero ([-0.1, +0.1], in the logit scale). This area indicates the values of the regression coefficients that we considered as equivalent to zero. We then computed the 95% posterior credible intervals (CrI) of the regression coefficients of interest. Finally, we calculated the proportion of the 95% CrI that fell inside the ROPE. This proportion indicates the probability that the true value of the coefficient is equivalent to zero. All analyses were performed in R environment [@rcoreteam2013language]. We used the tidyverse family of R packages [@wickham2019welcome] to process data and to generate figures. We used the `brms` R package [@burkner2017brms] using the `cmdstanr` back-end to the Stan probabilistic language [@carpenter2017stan] to estimate and compare the models (see Appendix 1 for mode details on the models).

@@

We collected data for a total of `r format(nrow(dataset$cat_eng) + nrow(dataset$spa_eng), big.mark = ",")` trials completed by `r n_participants$cat_eng + n_participants$spa_eng` unique participants. Of those trials, `r format(nrow(dataset$cat_eng), big.mark = ",")` were provided by `r n_participants$cat_eng` unique participants who listened to Catalan words, and `r format(nrow(dataset$spa_eng), big.mark = ",")` trials were provided by `r n_participants$spa_eng` unique participants who listened to Spanish words. We excluded trials in which participants did not enter any text (*n* = `r r_validity$cat_eng["blank"]`), in which a response in a language other than English was provided (e.g., `agua`, *n* = `r r_validity$cat_eng["language"]`), in which participants did not provide a whole word (e.g., `f`, *n* = `r r_validity$cat_eng["incomplete"]`), and in which participants added comments to the experimenter (e.g., `unsure`, *n* = `r r_validity$cat_eng["other"]`). In addition, we excluded data from participants that self-rated their oral and/or written skills in Catalan and Spanish, or any other second language as four or higher in a five-point scale (*n* = `r excluded_lang`), were diagnosed with a language (*n* = `r excluded_impairment`), or did not contribute more than 80% of valid trials (*n* = `r excluded_ntrials`).

@@

```{r fig-epreds-1}
#| label: fig-epreds-1
#| fig-height: 5
#| fig-width: 11
#| fig-cap: "Posterior model-predicted mean accuracy in Experiment 1. Predictions were generated from 4,000 posterior samples, extracted for different values of *CLPN* (0, 2, 4, 8, 12) and *Similarity* (1-100). Predictions are plotted separately for English participants translating Catalan words, and for English participants translating Spanish words. Lines indicate mean predictions, and intervals indicate 95%, 89%, 78%, 67%, and 50% credible intervals (CrI)."
get_epreds <- function(model, data, n = 1000, 
                       lv = seq(0, 1, length.out = 100), 
                       neigh_n_h = c(0, 2, 4, 8, 12),
                       ...) {

@@

On each trial, participants listened to the audio stimulus by clicking on the `PLAY` button. For comparability to the PsychoPy version, participants were only allowed to play the audio one time. Participants were explicitly told that they would be only allowed to listen once. The `PLAY` button vanished after one playthrough. Participants then had to answer three questions based on the audio they had heard on that trial. These questions were presented on the same page, directly below the audio player. They were first asked whether or not they knew the presented word (multiple choice---*yes*/*no*). Regardless of their answer on the first question, participants were asked what they thought the translation of the word was in English (or their best guess), and instructed to type their answer in the provided text box. Finally, they were asked to rate how confident they were in their answer on a scale of 0 to 7, where 7 was "very confident" and 0 was "not confident". There was no time limit on the response phase. All questions had to be answered to proceed to the next trial.
